### 2013-03-13 bwamem.tex
The unfinished version.

\documentclass{bioinfo}
\copyrightyear{2013}
\pubyear{2013}

\usepackage{natbib}
\bibliographystyle{apalike}

\DeclareMathOperator*{\argmax}{argmax}

\begin{document}
\firstpage{1}

\title{Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM}

\author[Li]{Heng Li$^{1,}$\footnote{to whom correspondence should be addressed}}

\address{$^1$Broad Institute, 7 Cambridge Center, Cambridge, MA 02142, USA}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
\editor{Associate Editor: XXXXXXX}
\maketitle

\begin{abstract}
\section{Summary:} BWA-MEM is a new alignment algorithm for aligning sequence
reads or long query sequences against a large reference genome such as human.
It automatically chooses between local and end-to-end alignments, supports
paired-end reads and performs split alignment. The algorithm is robust to
sequencing errors and applicable to a wide range of sequence lengths from 70bp
to a couple of megabases. For short-read mapping, BWA-MEM shows better
performance and alignment accuracy than many state-of-art read aligners to
date.

\section{Availability and implementation:} BWA-MEM is implemented as a
component of BWA, which is available at http://github.com/lh3/bwa.

\section{Contact:} hengli@broadinstitute.org
\end{abstract}

\section{Introduction}

Most short-reads mappers for next-generation sequencing (NGS) data were developed
when sequence reads were about 36bp in length. For short reads, it is reasonable
to require end-to-end alignment (i.e. every read base to be aligned to the
reference) and to only report hits within certain hamming or edit distance.
However, with emerging technologies and improved chemistry, NGS reads are not
short any more, which poses new challenges to read alignment. For longer reads,
it becomes more important to allow long gaps under the affine-gap penalty and
to report multiple non-overlapping local hits potentially caused by structural
variations or misassemblies in the reference genome. Many short-read alignment
algorithms are not applicable or not preferred for mapping longer reads. At the
same time, although several mature algorithms exist for aligning capillary
sequence reads, they are slow and lack features for analyzing large-scale NGS
data. Fast moving NGS technologies keep pressing for the development of new
alignment algorithms.

In this background, a few long-read alignment algorithms, including BWA-SW,
Bowtie2, Cushaw2 and GEM, have been developed. However, they all have some
weakness. BWA-SW is slower than bowtie2 at a comparable accuracy and less
accurate then Cushaw2 at a comparable speed (see Results). Bowtie2 and Cushaw2
do not work with very long query sequences. These three aligners do not fully
use the paired-end information to improve alignments in hard regions, either.
While GEM is both fast and accurate, it mandates end-to-end alignment and does
not perform affine-gap alignment, which may limit its uses for long-read
alignment. A fast, accurate and fully featured aligner accepting sequences
with a wide range of lengths is still lacking. This motivates us to explore
a new algorithm, BWA-MEM.

\begin{methods}

\section{Methods}

\subsection{Aligning single query sequence}

\subsubsection{Seeding and re-seeding} BWA-MEM follows the canonical
seed-and-extend paradigm. It initially seeds an alignment with supermaximal
exact matches (SMEMs) using an algorithm we found previously. This algorithm
essentially finds at each query position the longest exact match covering the
position. Suppose we have a SMEM of length $l$ with $k$ occurrences in the
reference genome. If $l$ is too large (over 28bp by default), we then re-seed
with exact matches that cover the middle base of the SMEM and occur at least
$k+1$ times in the genome, which can be done by requiring a minimum occurrence
in the original SMEM algorithm. Re-seeding reduces the chance of missing a seed
from the true alignment.

\subsubsection{Chaining and chain filtering} We call a group of seeds that are
colinear and close to each other as a \emph{chain}. While seeding, we chain
the seeds with a naive algorithm. We then filter out short chains that are
largely contained in a long chain and is much worse than the long chain (by
default, both 50\% and 38bp shorter than the long chain). Note that chain
filtering aims to reduce unsuccessful seed extension at a later step. The
chains detected here do not need to be accurate. We may implicitly split or
merge chains later.

\subsubsection{Seed extension} We rank the seeds by the length of the chains
the seeds belong to and then by the length of the seeds in each chain. We go
through each seed, from the best ranked to the worst, and drop the seed if it
is already ontained in an alignment found before, or use a banded
affine-gap-penalty dynamic programming to extend the seed in either direction
if the seed potentially leads to a new alignment. Up to here, we find all
the potential alignments of a query sequence.

BWA-MEM's deed extension differs from the standard seed extension in two
aspects. Firstly, suppose at a certain extension step we come to reference
position $x$ with the best extension score achieved at query position $y$.
BWA-MEM will stop extension if the score at $(x,y)$ is smaller than the best
extension score observed so far by $Z+|x-y|\times p_{\rm gapExt}$, where
$p_{\rm gapExt}$ is the gap extension penalty. This heuristic avoids extending
through a poorly aligned region with good flanking alignment. It is similar
to the X-dropoff heuristic in BLAST, but does not penalize long gaps in
one of the reference or the query sequences.

Secondly, while extending a seed, BWA-MEM tries to keep track of the best
extension score reaching the end of the query sequence. If the difference
between the best score reaching the end and the best local alignment score is
below a threshold, which is called clipping penalty, the local alignment will
be rejected even if it has a higher score. BWA-MEM uses this strategy to
automatically choose between local and end-to-end alignments. It may allow true
substitutions towards the end of a read and thus reduces reference bias, while
avoids introducing excessive mismatches and gaps which may happen when we force
an end-to-end alignment for a read bridging a break point associated with a
long insertion or deletion (indel) or a structural variation or a misassembly.

\subsection{Paired-end mapping}

\subsubsection{Estimating insert size distribution and rescuing missing hits}
Like BWA, BWA-MEM processes a batch of reads at a time.  Given paired-end data,
it estimates the mean $\mu$ and the variance $\sigma^2$ of the insert size
distribution from reliable single-end alignments. For the top 100 hits (by
default) of each end, BWA-MEM performs the Smith-Waterman alignment (SW) for
its mate within a window $[\mu-4\sigma,\mu+4\sigma]$ from the hit. The second
best alignment score is recorded to detect potential mismapping in a long
tandem repeat.  Hits found both from the single-sequence alignment and SW
rescuing will be fed to the next step for pairing.

\subsubsection{Pairing} Given the $i$-th hit for the first read and $j$-th hit
for the second, BWA-MEM computes their distance $d_{ij}$ if the two hits are in
the right orientation, or sets $d_{ij}$ to infinity otherwise. It scores the
pair $(i,j)$ as $S_{ij}=S_i+S_j-\min\{-a\log_4 P(d_{ij}),U\}$, where $S_i$ and
$S_j$ are the SW score of the two reads in the pair, respectively, $a$ is the
matching score and $P(d)$ gives the probability of observing an insert size
larger than $d$ assuming a normal distribution. `$\log_4$' arises when we
relate SW score to a probabilistic alignment. $U$ is the unpairing penalty,
which controls whether pairing is preferred: if $d_{ij}$ is small enough (such
that $-a\log_4 P(d_{ij})<U$), BWA-MEM prefers to pair the two ends; otherwise
it prefers the unpaired alignments. BWA-MEM chooses the pair $(i,j)$ that
maximizes $S_{ij}$ as the final alignment for both ends.

This pairing strategy jointly considers the alignment scores of both ends,
the insert size and the possibility that they may come from a chimeric read
pair. It prefers pairing the two ends but does not aggressively force two ends
into a read pair.

\end{methods}

\paragraph{Funding\textcolon} NIH 1U01HG005208-01.
\bibliography{bwamem}

\end{document}

### 2013-03-14 bwamem.tex
Added brief results. More works needed of course.

@@
 to a couple of megabases. For short-read mapping, BWA-MEM shows better
-performance and alignment accuracy than many state-of-art read aligners to
-date.
+performance than many state-of-art read aligners to date.
 
@@
 
-In this background, a few long-read alignment algorithms, including BWA-SW,
-Bowtie2, Cushaw2 and GEM, have been developed. However, they all have some
+In this background, a few long-read alignment algorithms, including BWA-SW~\citep{Li:2010fk},
+Bowtie2~\citep{Langmead:2012fk}, Cushaw2~\citep{Liu:2012uq} and GEM~\citep{Marco-Sola:2012kx}, have been developed. However, they all have some
 weakness. BWA-SW is slower than bowtie2 at a comparable accuracy and less
@@
 with a wide range of lengths is still lacking. This motivates us to explore
-a new algorithm, BWA-MEM.
+a new algorithm.
 
@@
 seed-and-extend paradigm. It initially seeds an alignment with supermaximal
-exact matches (SMEMs) using an algorithm we found previously. This algorithm
+exact matches (SMEMs) using an algorithm we found previously~\citep[Algorithm 5]{Li:2012fk}. This algorithm
 essentially finds at each query position the longest exact match covering the
@@
 chains detected here do not need to be accurate. We may implicitly split or
-merge chains later.
+merge chains during seed extension.
 
@@
 through a poorly aligned region with good flanking alignment. It is similar
-to the X-dropoff heuristic in BLAST, but does not penalize long gaps in
+to the X-dropoff heuristic in BLAST~\citep{Altschul:1997vn}, but does not penalize long gaps in
 one of the reference or the query sequences.
@@
 long insertion or deletion (indel) or a structural variation or a misassembly.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[width=0.4\textwidth]{eval/alnroc-se}\\
+\includegraphics[width=0.4\textwidth]{eval/alnroc-pe}
+\caption{Percent mapped reads as a function of the false alignment rate under
+different mapping quality cutoff. $10^6$ pairs of 101bp reads are simulated
+from the human reference genome using wgsim (http://bit.ly/wgsim2) with 1.5\%
+substitution errors and 0.2\% indel variants. The insert size follows a normal
+distribution $N(500,50^2)$. The reads are aligned back to the genome either as
+single end (top panel) or as paired end (bottom panel). GEM is configured to
+allow gaps and output suboptimal alignments (option --e5 --m5 --s1). Its
+mapping quality is estimated with a BWA-like algorithm. Other mappers are run
+essentially with the default setting except specifying the insert size
+distribution. CPU timeU is shown in the parantheses.} \end{figure}\label{fig:eval}.
 
@@
 
+\section{Results}
+
+We evaluated the performance of BWA-MEM on simulated data together with
+GEM, Bowtie2, Cushaw2, BWA-SW, BWA~\citep{Li:2009uq} and NovoAlign
+(http://novocraft.com) (Figure~\ref{fig:eval}). We were unable to obtain high
+accuracy for paired-end data with GEM, so we did not show it result. In this
+evaluation, NovoAlign is the most accurate. BWA-MEM is only second to NovoAlign
+in accuracy and among the fastest in speed.
+
+\section{Conclusion}
+
+BWA-MEM is a fast and accurate alignment algorithm for sequence reads. It is
+one of the few aligners that work with both short reads and long sequences up
+to a few megabases.
+
+\section{Acknowledgements}
 \paragraph{Funding\textcolon} NIH 1U01HG005208-01.

### 2013-03-14 bwamem.tex
the first version

@@
 \subsubsection{Seed extension} We rank the seeds by the length of the chains
-the seeds belong to and then by the length of the seeds in each chain. We go
-through each seed, from the best ranked to the worst, and drop the seed if it
-is already ontained in an alignment found before, or use a banded
-affine-gap-penalty dynamic programming to extend the seed in either direction
-if the seed potentially leads to a new alignment. Up to here, we find all
-the potential alignments of a query sequence.
+the seeds belong to and then by the length of the seeds in each chain. We start
+from the highest ranked seed and for each seed, we drop the seed if it is
+already contained in an alignment found before, or extend with a
+affine-gap-penalty dynamic programming (DP) otherwise. We will find all the
+potential alignments of a query sequence after this step.
 
-BWA-MEM's deed extension differs from the standard seed extension in two
+BWA-MEM's seed extension differs from the standard seed extension in two
 aspects. Firstly, suppose at a certain extension step we come to reference
@@
 extension score observed so far by $Z+|x-y|\times p_{\rm gapExt}$, where
-$p_{\rm gapExt}$ is the gap extension penalty. This heuristic avoids extending
+$p_{\rm gapExt}$ is the gap extension penalty and $Z$ is an arbitrary value. This heuristic avoids extending
 through a poorly aligned region with good flanking alignment. It is similar
@@
 distribution $N(500,50^2)$. The reads are aligned back to the genome either as
-single end (top panel) or as paired end (bottom panel). GEM is configured to
-allow gaps and output suboptimal alignments (option --e5 --m5 --s1). Its
+single end (SE; top panel) or as paired end (PE; bottom panel). GEM is configured to
+allow up to 5 gaps and to output suboptimal alignments (option `--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). Its
 mapping quality is estimated with a BWA-like algorithm. Other mappers are run
 essentially with the default setting except specifying the insert size
-distribution. CPU timeU is shown in the parantheses.} \end{figure}\label{fig:eval}.
+distribution. The run time on a single CPU core is shown in the parantheses.}\label{fig:eval}
+\end{figure}.
 
@@
 larger than $d$ assuming a normal distribution. `$\log_4$' arises when we
-relate SW score to a probabilistic alignment. $U$ is the unpairing penalty,
-which controls whether pairing is preferred: if $d_{ij}$ is small enough (such
-that $-a\log_4 P(d_{ij})<U$), BWA-MEM prefers to pair the two ends; otherwise
+relate SW score to a probabilistic model. $U$ is the chimeric penalty,
+which controls preference of pairing: if $d_{ij}$ is small enough such
+that $-a\log_4 P(d_{ij})<U$, BWA-MEM prefers to pair the two ends; otherwise
 it prefers the unpaired alignments. BWA-MEM chooses the pair $(i,j)$ that
 maximizes $S_{ij}$ as the final alignment for both ends.
-
 This pairing strategy jointly considers the alignment scores of both ends,
-the insert size and the possibility that they may come from a chimeric read
-pair. It prefers pairing the two ends but does not aggressively force two ends
-into a read pair.
+the insert size and the possibility of chimeric read pairs.
 
@@
 
-\section{Results}
+\section{Results and Discussions}
 
@@
 GEM, Bowtie2, Cushaw2, BWA-SW, BWA~\citep{Li:2009uq} and NovoAlign
-(http://novocraft.com) (Figure~\ref{fig:eval}). We were unable to obtain high
-accuracy for paired-end data with GEM, so we did not show it result. In this
-evaluation, NovoAlign is the most accurate. BWA-MEM is only second to NovoAlign
-in accuracy and among the fastest in speed.
+(http://novocraft.com) (Figure~\ref{fig:eval}). On accuracy, NovoAlign
+is the most accurate. BWA-MEM is close to NovoAlign for PE reads
+and is comparable to GEM and Cushaw2 for SE. On speed, BWA-MEM is similar to
+GEM and Bowtie2 for this data set, but is about 6 times as fast as Bowtie2
+for a 650bp long-read data set.
 
-\section{Conclusion}
+To evaluate the performance for even long query sequences, we aligned the {\it E. coli}
+K-12 MG1655 substrain (AC:NC\_000913) against the 536 strain (AC:NC\_008253) with 
+both BWA-MEM and nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds, giving
+105,505 substitution differences between the strains. BWA-MEM is slower.
+It finished the alignment in 131 seconds, reporting 104,321 substitutions of which 102,241
+overlapping with the nucmer output. Manual examination of the
+substitutions unique to each aligner reveals that most of them fall in short regions
+with very high divergence. It is unclear which aligner is better. Although
+Nucmer is faster in the evaluation, only BWA-MEM scales to mammalian genomes.
 
-BWA-MEM is a fast and accurate alignment algorithm for sequence reads. It is
-one of the few aligners that work with both short reads and long sequences up
-to a few megabases.
+BWA-MEM is a fast and accurate alignment algorithm for sequence reads and is
+one of the few aligners that work efficiently with both short reads and long sequences up
+to a few megabases. It is still possible to greatly accelerate BWA-MEM for long
+sequences by using SSE2-powered banded DP and by restricting DP to regions not
+covered by a long seed.
 

### 2013-03-14 bwamem.tex
improve wording, a little bit

@@
 \bibliographystyle{apalike}
-
-\DeclareMathOperator*{\argmax}{argmax}
 
@@
 
-Most short-reads mappers for next-generation sequencing (NGS) data were developed
-when sequence reads were about 36bp in length. For short reads, it is reasonable
-to require end-to-end alignment (i.e. every read base to be aligned to the
-reference) and to only report hits within certain hamming or edit distance.
-However, with emerging technologies and improved chemistry, NGS reads are not
-short any more, which poses new challenges to read alignment. For longer reads,
-it becomes more important to allow long gaps under the affine-gap penalty and
-to report multiple non-overlapping local hits potentially caused by structural
-variations or misassemblies in the reference genome. Many short-read alignment
-algorithms are not applicable or not preferred for mapping longer reads. At the
-same time, although several mature algorithms exist for aligning capillary
-sequence reads, they are slow and lack features for analyzing large-scale NGS
-data. Fast moving NGS technologies keep pressing for the development of new
-alignment algorithms.
+Most short-reads mappers for next-generation sequencing (NGS) data were
+developed when sequence reads were about 36bp in length. For short reads, it is
+reasonable to require end-to-end alignment (i.e. every read base to be aligned
+to the reference) and to only report hits within certain hamming or edit
+distance.  However, with emerging technologies and improved chemistry, NGS
+reads are not short any more, which poses new challenges to read alignment. For
+longer reads, it becomes more important to allow long gaps under the affine-gap
+penalty and to report multiple non-overlapping local hits potentially caused by
+structural variations or misassemblies in the reference genome. Many short-read
+alignment algorithms are not applicable or not preferred for mapping longer
+reads. At the same time, although several mature algorithms exist for aligning
+capillary sequence reads, they are slow and lack features for analyzing
+large-scale NGS data. Fast moving NGS technologies keep pressing for the
+development of new alignment algorithms.
 
-In this background, a few long-read alignment algorithms, including BWA-SW~\citep{Li:2010fk},
-Bowtie2~\citep{Langmead:2012fk}, Cushaw2~\citep{Liu:2012uq} and GEM~\citep{Marco-Sola:2012kx}, have been developed. However, they all have some
-weakness. BWA-SW is slower than bowtie2 at a comparable accuracy and less
-accurate then Cushaw2 at a comparable speed (see Results). Bowtie2 and Cushaw2
-do not work with very long query sequences. These three aligners do not fully
-use the paired-end information to improve alignments in hard regions, either.
-While GEM is both fast and accurate, it mandates end-to-end alignment and does
-not perform affine-gap alignment, which may limit its uses for long-read
-alignment. A fast, accurate and fully featured aligner accepting sequences
-with a wide range of lengths is still lacking. This motivates us to explore
-a new algorithm.
+In this background, a few long-read alignment algorithms, including
+BWA-SW~\citep{Li:2010fk}, Bowtie2~\citep{Langmead:2012fk},
+Cushaw2~\citep{Liu:2012uq} and GEM~\citep{Marco-Sola:2012kx}, have been
+developed. However, they all have some weakness. BWA-SW is slower than bowtie2
+at a comparable accuracy and less accurate then Cushaw2 at a comparable speed
+(see Results). Bowtie2 and Cushaw2 do not work with very long query sequences.
+These three aligners do not fully use the paired-end information to improve
+alignments in hard regions, either.  While GEM is both fast and accurate, it
+mandates end-to-end alignment and does not perform affine-gap alignment, which
+may limit its uses for long-read alignment. A fast, accurate and fully featured
+aligner accepting sequences with a wide range of lengths is still lacking. This
+motivates us to explore a new algorithm.
 
+\vspace*{-1em}
 \begin{methods}
@@
 
-\subsection{Aligning single query sequence}
+\subsection{Aligning a single query sequence}
 
@@
 seed-and-extend paradigm. It initially seeds an alignment with supermaximal
-exact matches (SMEMs) using an algorithm we found previously~\citep[Algorithm 5]{Li:2012fk}. This algorithm
-essentially finds at each query position the longest exact match covering the
-position. Suppose we have a SMEM of length $l$ with $k$ occurrences in the
+exact matches (SMEMs) using an algorithm we found previously~\citep[Algorithm
+5]{Li:2012fk}, which essentially finds at each query position the longest exact
+match covering the position. However, occasionally the true alignment may not
+contain any SMEMs.  To reduce mismappings caused by missing seeds, we introduce
+re-seeding. Suppose we have a SMEM of length $l$ with $k$ occurrences in the
 reference genome. If $l$ is too large (over 28bp by default), we then re-seed
 with exact matches that cover the middle base of the SMEM and occur at least
-$k+1$ times in the genome, which can be done by requiring a minimum occurrence
-in the original SMEM algorithm. Re-seeding reduces the chance of missing a seed
-from the true alignment.
+$k+1$ times in the genome. Such seeds can be found by requiring a minimum
+occurrence in the original SMEM algorithm.
 
 \subsubsection{Chaining and chain filtering} We call a group of seeds that are
-colinear and close to each other as a \emph{chain}. While seeding, we chain
-the seeds with a naive algorithm. We then filter out short chains that are
-largely contained in a long chain and is much worse than the long chain (by
-default, both 50\% and 38bp shorter than the long chain). Note that chain
-filtering aims to reduce unsuccessful seed extension at a later step. The
-chains detected here do not need to be accurate. We may implicitly split or
-merge chains during seed extension.
+colinear and close to each other as a \emph{chain}. We chain the seeds while
+seeding. After that, we filter out short chains that are largely contained in a
+long chain and are much worse than the long chain (by default, both 50\% and
+38bp shorter than the long chain). Note that chain filtering aims to reduce
+unsuccessful seed extension at a later step. Each chain may not always
+correspond to a final hit. The chains detected here do not need to be accurate.
 
-\subsubsection{Seed extension} We rank the seeds by the length of the chains
-the seeds belong to and then by the length of the seeds in each chain. We start
-from the highest ranked seed and for each seed, we drop the seed if it is
-already contained in an alignment found before, or extend with a
-affine-gap-penalty dynamic programming (DP) otherwise. We will find all the
-potential alignments of a query sequence after this step.
+\subsubsection{Seed extension} We rank a seed by the length of the chain it
+belongs to and then by the seed length in the chain. For each seed in the
+ranked list, from the best to the worst, we drop the seed if it is already
+contained in an alignment found before, or extend the seed with a banded
+affine-gap-penalty dynamic programming (DP) if it potentially leads to a new
+alignment.
 
@@
 position $x$ with the best extension score achieved at query position $y$.
-BWA-MEM will stop extension if the score at $(x,y)$ is smaller than the best
-extension score observed so far by $Z+|x-y|\times p_{\rm gapExt}$, where
-$p_{\rm gapExt}$ is the gap extension penalty and $Z$ is an arbitrary value. This heuristic avoids extending
-through a poorly aligned region with good flanking alignment. It is similar
-to the X-dropoff heuristic in BLAST~\citep{Altschul:1997vn}, but does not penalize long gaps in
-one of the reference or the query sequences.
+BWA-MEM will stop extension if the difference between the best extension score
+and the score at $(x,y)$ is larger than $Z+|x-y|\times p_{\rm gapExt}$, where
+$p_{\rm gapExt}$ is the gap extension penalty and $Z$ is an arbitrary cutoff.
+This heuristic avoids extension through a poorly aligned region with good
+flanking alignment. It is similar to the X-dropoff heuristic in
+BLAST~\citep{Altschul:1997vn}, but does not penalize long gaps in one of the
+reference or the query sequences.
 
@@
 between the best score reaching the end and the best local alignment score is
-below a threshold, which is called clipping penalty, the local alignment will
-be rejected even if it has a higher score. BWA-MEM uses this strategy to
-automatically choose between local and end-to-end alignments. It may allow true
-substitutions towards the end of a read and thus reduces reference bias, while
-avoids introducing excessive mismatches and gaps which may happen when we force
-an end-to-end alignment for a read bridging a break point associated with a
-long insertion or deletion (indel) or a structural variation or a misassembly.
+below a threshold, the local alignment will be rejected even if it has a higher
+score. BWA-MEM uses this strategy to automatically choose between local and
+end-to-end alignments. It may allow true substitutions towards the end of a
+read and thus reduces reference bias, while avoids introducing excessive
+mismatches and gaps which may happen when we force an end-to-end alignment for
+a read bridging a break point associated with a long insertion or deletion
+(indel) or a structural variation or a misassembly.
 
@@
 distribution $N(500,50^2)$. The reads are aligned back to the genome either as
-single end (SE; top panel) or as paired end (PE; bottom panel). GEM is configured to
-allow up to 5 gaps and to output suboptimal alignments (option `--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). Its
-mapping quality is estimated with a BWA-like algorithm. Other mappers are run
-essentially with the default setting except specifying the insert size
-distribution. The run time on a single CPU core is shown in the parantheses.}\label{fig:eval}
+single end (SE; top panel) or as paired end (PE; bottom panel). GEM is
+configured to allow up to 5 gaps and to output suboptimal alignments (option
+`--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). Its mapping quality
+is estimated with a BWA-like algorithm. Other mappers are run essentially with
+the default setting except specifying the insert size distribution. The run
+time on a single CPU core is shown in the parantheses.}\label{fig:eval}
 \end{figure}.
@@
 
-\subsubsection{Estimating insert size distribution and rescuing missing hits}
-Like BWA, BWA-MEM processes a batch of reads at a time.  Given paired-end data,
-it estimates the mean $\mu$ and the variance $\sigma^2$ of the insert size
-distribution from reliable single-end alignments. For the top 100 hits (by
-default) of each end, BWA-MEM performs the Smith-Waterman alignment (SW) for
-its mate within a window $[\mu-4\sigma,\mu+4\sigma]$ from the hit. The second
-best alignment score is recorded to detect potential mismapping in a long
-tandem repeat.  Hits found both from the single-sequence alignment and SW
-rescuing will be fed to the next step for pairing.
+\subsubsection{Rescuing missing hits}
+Like BWA~\citep{Li:2009uq}, BWA-MEM processes a batch of reads at a time.
+For each batch, it estimates the mean $\mu$ and the variance $\sigma^2$
+of the insert size distribution from reliable single-end hits. For the
+top 100 hits (by default) of each end, BWA-MEM performs SSE2-based
+Smith-Waterman alignment (SW; \citealt{Farrar:2007hs}) for the mate within a
+window $[\mu-4\sigma,\mu+4\sigma]$ from each hit. The second best SW score is
+recorded to detect potential mismapping in a long tandem repeat. Hits found
+from both the single-sequence alignment and SW rescuing will be used for
+pairing. It is worth noting that BWA and BWA-SW only performs SW for the
+best unique hit, but BWA-MEM also does SW for repetitive hits.
 
@@
 the right orientation, or sets $d_{ij}$ to infinity otherwise. It scores the
-pair $(i,j)$ as $S_{ij}=S_i+S_j-\min\{-a\log_4 P(d_{ij}),U\}$, where $S_i$ and
-$S_j$ are the SW score of the two reads in the pair, respectively, $a$ is the
-matching score and $P(d)$ gives the probability of observing an insert size
-larger than $d$ assuming a normal distribution. `$\log_4$' arises when we
-relate SW score to a probabilistic model. $U$ is the chimeric penalty,
-which controls preference of pairing: if $d_{ij}$ is small enough such
-that $-a\log_4 P(d_{ij})<U$, BWA-MEM prefers to pair the two ends; otherwise
-it prefers the unpaired alignments. BWA-MEM chooses the pair $(i,j)$ that
-maximizes $S_{ij}$ as the final alignment for both ends.
-This pairing strategy jointly considers the alignment scores of both ends,
-the insert size and the possibility of chimeric read pairs.
+pair $(i,j)$ with $S_{ij}=S_i+S_j-\min\{-a\log_4 P(d_{ij}),U\}$, where $S_i$
+and $S_j$ are the SW score of the two hits, respectively, $a$ is the matching
+score and $P(d)$ gives the probability of observing an insert size larger than
+$d$ assuming a normal distribution. `$\log_4$' arises when we interpret SW
+score as odds ratio~\citep{Durbin:1998uq}. $U$ is a threshold that controls
+pairing: if $d_{ij}$ is small enough such that $-a\log_4 P(d_{ij})<U$, BWA-MEM
+prefers to pair the two ends; otherwise it prefers the unpaired alignments.
+BWA-MEM chooses the pair $(i,j)$ that maximizes $S_{ij}$ as the final
+alignments for both ends.  This pairing strategy jointly considers the
+alignment scores, the insert size and the possibility of chimeric read pairs.
 
@@
 We evaluated the performance of BWA-MEM on simulated data together with
-GEM, Bowtie2, Cushaw2, BWA-SW, BWA~\citep{Li:2009uq} and NovoAlign
-(http://novocraft.com) (Figure~\ref{fig:eval}). On accuracy, NovoAlign
-is the most accurate. BWA-MEM is close to NovoAlign for PE reads
-and is comparable to GEM and Cushaw2 for SE. On speed, BWA-MEM is similar to
-GEM and Bowtie2 for this data set, but is about 6 times as fast as Bowtie2
-for a 650bp long-read data set.
+NovoAlign (http://novocraft.com), GEM, Bowtie2, Cushaw2, BWA-SW and BWA
+(Figure~\ref{fig:eval}). On accuracy, NovoAlign is the best. BWA-MEM is close
+to NovoAlign for PE reads and is comparable to GEM and Cushaw2 for SE.  On
+speed, BWA-MEM is similar to GEM and Bowtie2 for this data set, but is about
+6 times as fast as Bowtie2 for a 650bp long-read data set.
 
-To evaluate the performance for even long query sequences, we aligned the {\it E. coli}
-K-12 MG1655 substrain (AC:NC\_000913) against the 536 strain (AC:NC\_008253) with 
-both BWA-MEM and nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds, giving
-105,505 substitution differences between the strains. BWA-MEM is slower.
-It finished the alignment in 131 seconds, reporting 104,321 substitutions of which 102,241
-overlapping with the nucmer output. Manual examination of the
-substitutions unique to each aligner reveals that most of them fall in short regions
-with very high divergence. It is unclear which aligner is better. Although
-Nucmer is faster in the evaluation, only BWA-MEM scales to mammalian genomes.
+To evaluate the performance for even longer query sequences, we aligned the
+{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
+the 536 strain (AC:NC\_008253) with both BWA-MEM and
+nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
+giving 105,505 substitution differences between the strains. BWA-MEM is slower.
+It finished the alignment in 131 seconds, reporting 104,321 substitutions of
+which 102,241 overlapping with the nucmer output. Manual examination of the
+substitutions unique to each aligner reveals that most of them fall in short
+regions with very high divergence. It is unclear which aligner is better. Note
+that although Nucmer is faster in the evaluation, only BWA-MEM scales well to
+large genomes.
 
-BWA-MEM is a fast and accurate alignment algorithm for sequence reads and is
-one of the few aligners that work efficiently with both short reads and long sequences up
-to a few megabases. It is still possible to greatly accelerate BWA-MEM for long
-sequences by using SSE2-powered banded DP and by restricting DP to regions not
-covered by a long seed.
+BWA-MEM is a fast and accurate aligner for sequence reads and is one of the few
+that work well for both short reads and long sequences up to a few megabases.
+Technically, it is possible to make BWA-MEM even faster for long sequences by
+using SSE2-based banded DP and by restricting DP to regions not covered by a
+long exact match.
 

### 2013-03-14 bwamem.tex
a little more wording improvement

@@
 
-\author[Li]{Heng Li$^{1,}$\footnote{to whom correspondence should be addressed}}
+\author[Li]{Heng Li}
 
-\address{$^1$Broad Institute, 7 Cambridge Center, Cambridge, MA 02142, USA}
+\address{Broad Institute of Harvard and MIT, 7 Cambridge Center, Cambridge, MA 02142, USA}
 
@@
 sequencing errors and applicable to a wide range of sequence lengths from 70bp
-to a couple of megabases. For short-read mapping, BWA-MEM shows better
-performance than many state-of-art read aligners to date.
+to a few megabases. For short-read mapping, BWA-MEM shows better performance
+than many state-of-art read aligners to date.
 
@@
 
-\vspace*{-1em}
 \begin{methods}
@@
 re-seeding. Suppose we have a SMEM of length $l$ with $k$ occurrences in the
-reference genome. If $l$ is too large (over 28bp by default), we then re-seed
-with exact matches that cover the middle base of the SMEM and occur at least
-$k+1$ times in the genome. Such seeds can be found by requiring a minimum
-occurrence in the original SMEM algorithm.
+reference genome. If $l$ is too large (over 28bp by default), we re-seed
+with the longest exact matches that cover the middle base of the SMEM and occur
+at least $k+1$ times in the genome. Such seeds can be found by requiring a
+minimum occurrence in the original SMEM algorithm.
 
@@
 long chain and are much worse than the long chain (by default, both 50\% and
-38bp shorter than the long chain). Note that chain filtering aims to reduce
-unsuccessful seed extension at a later step. Each chain may not always
-correspond to a final hit. The chains detected here do not need to be accurate.
+38bp shorter than the long chain). Chain filtering aims to reduce unsuccessful
+seed extension at a later step. Each chain may not always correspond to a final
+hit. Chains detected here do not need to be accurate.
 
 \subsubsection{Seed extension} We rank a seed by the length of the chain it
-belongs to and then by the seed length in the chain. For each seed in the
-ranked list, from the best to the worst, we drop the seed if it is already
-contained in an alignment found before, or extend the seed with a banded
-affine-gap-penalty dynamic programming (DP) if it potentially leads to a new
-alignment.
+belongs to and then by the seed length. For each seed in the ranked list, from
+the best to the worst, we drop the seed if it is already contained in an
+alignment found before, or extend the seed with a banded affine-gap-penalty
+dynamic programming (DP) if it potentially leads to a new alignment.
 
@@
 score. BWA-MEM uses this strategy to automatically choose between local and
-end-to-end alignments. It may allow true substitutions towards the end of a
-read and thus reduces reference bias, while avoids introducing excessive
-mismatches and gaps which may happen when we force an end-to-end alignment for
-a read bridging a break point associated with a long insertion or deletion
-(indel) or a structural variation or a misassembly.
+end-to-end alignments. It may align through true variations towards the end of
+a read and thus reduces reference bias, while avoids introducing excessive
+mismatches and gaps which may happen when we force a chimeric read into an
+end-to-end alignment.
 
@@
 is estimated with a BWA-like algorithm. Other mappers are run essentially with
-the default setting except specifying the insert size distribution. The run
-time on a single CPU core is shown in the parantheses.}\label{fig:eval}
+the default setting except for specifying the insert size distribution. The CPU
+time in seconds is reported in the parantheses.}\label{fig:eval}
 \end{figure}.
@@
 using SSE2-based banded DP and by restricting DP to regions not covered by a
-long exact match.
+long exact match. Seeding is the bottleneck for short sequences, while banded
+DP is the bottleneck for long sequences.
 

### 2013-03-15 bwamem.tex
put figures in the base directory, not in eval/

@@
 to a few megabases. For short-read mapping, BWA-MEM shows better performance
-than many state-of-art read aligners to date.
+than several state-of-art read aligners to date.
 
@@
 colinear and close to each other as a \emph{chain}. We chain the seeds while
-seeding. After that, we filter out short chains that are largely contained in a
-long chain and are much worse than the long chain (by default, both 50\% and
-38bp shorter than the long chain). Chain filtering aims to reduce unsuccessful
-seed extension at a later step. Each chain may not always correspond to a final
-hit. Chains detected here do not need to be accurate.
+seeding. After chains are collected, we filter out short chains that are
+largely contained in a long chain and are much worse than the long chain (by
+default, both 50\% and 38bp shorter than the long chain). Chain filtering aims
+to reduce unsuccessful seed extension at a later step. Each chain may not
+always correspond to a final hit. Chains detected here do not need to be
+accurate.
 
@@
 \centering
-\includegraphics[width=0.4\textwidth]{eval/alnroc-se}\\
-\includegraphics[width=0.4\textwidth]{eval/alnroc-pe}
+\includegraphics[width=0.4\textwidth]{alnroc-se}\\
+\includegraphics[width=0.4\textwidth]{alnroc-pe}
 \caption{Percent mapped reads as a function of the false alignment rate under
@@
 configured to allow up to 5 gaps and to output suboptimal alignments (option
-`--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). Its mapping quality
-is estimated with a BWA-like algorithm. Other mappers are run essentially with
+`--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). GEM does not compute
+mapping quality. Its mapping quality is estimated with a BWA-like algorithm
+with suboptimal alignments available. Other mappers are run essentially with
 the default setting except for specifying the insert size distribution. The CPU
-time in seconds is reported in the parantheses.}\label{fig:eval}
+time in seconds is shown in the parantheses.}\label{fig:eval}
 \end{figure}.
@@
 regions with very high divergence. It is unclear which aligner is better. Note
-that although Nucmer is faster in the evaluation, only BWA-MEM scales well to
+that although nucmer is faster in the evaluation, only BWA-MEM scales well to
 large genomes.

### 2013-04-27 bwamem.tex
one extra "." by mistake
updated evaluation

@@
 \section{Summary:} BWA-MEM is a new alignment algorithm for aligning sequence
-reads or long query sequences against a large reference genome such as human.
+reads or assembly contigs against a large reference genome such as human.
 It automatically chooses between local and end-to-end alignments, supports
-paired-end reads and performs split alignment. The algorithm is robust to
+paired-end reads and reports multiple local hits. The algorithm is robust to
 sequencing errors and applicable to a wide range of sequence lengths from 70bp
-to a few megabases. For short-read mapping, BWA-MEM shows better performance
-than several state-of-art read aligners to date.
+to a few megabases. For mapping 100bp sequences, BWA-MEM shows better
+performance than several state-of-art read aligners to date.
 
@@
 Most short-reads mappers for next-generation sequencing (NGS) data were
-developed when sequence reads were about 36bp in length. For short reads, it is
+developed when sequence reads were about 36bp in length. For 36bp reads, it is
 reasonable to require end-to-end alignment (i.e. every read base to be aligned
@@
 reads are not short any more, which poses new challenges to read alignment. For
-longer reads, it becomes more important to allow long gaps under the affine-gap
+100bp or longer reads, it becomes more important to allow long gaps under the affine-gap
 penalty and to report multiple non-overlapping local hits potentially caused by
@@
 developed. However, they all have some weakness. BWA-SW is slower than bowtie2
-at a comparable accuracy and less accurate then Cushaw2 at a comparable speed
-(see Results). Bowtie2 and Cushaw2 do not work with very long query sequences.
-These three aligners do not fully use the paired-end information to improve
-alignments in hard regions, either.  While GEM is both fast and accurate, it
+for 100bp reads at a comparable accuracy and less accurate then Cushaw2 at a
+comparable speed. Bowtie2 and Cushaw2 are slower for 600bp reads (see Results).
+While GEM is both fast and accurate for up to approximately 1000bp reads, it
 mandates end-to-end alignment and does not perform affine-gap alignment, which
-may limit its uses for long-read alignment. A fast, accurate and fully featured
-aligner accepting sequences with a wide range of lengths is still lacking. This
-motivates us to explore a new algorithm.
+limits its uses for long-read alignment. Even for typical sequence reads ranged
+from 100bp to 1000bp in length, no mappers work well across the full spectrum.
+At the same time, the increasing read length not only calls for new alignment
+algorithms, but also opens the opportunity to de novo assembly which typically
+results in contigs ranged from several hundred base pairs to a few megabases.
+Very few algorithms are able to align sequences of such variable lengths
+at high accuracy and to be robust to translocations in the assembly. All these
+concerns motivated us to explore a new alignment algorithm.
+
+\vspace{-1em}
 
@@
 \subsubsection{Chaining and chain filtering} We call a group of seeds that are
-colinear and close to each other as a \emph{chain}. We chain the seeds while
-seeding. After chains are collected, we filter out short chains that are
-largely contained in a long chain and are much worse than the long chain (by
-default, both 50\% and 38bp shorter than the long chain). Chain filtering aims
-to reduce unsuccessful seed extension at a later step. Each chain may not
-always correspond to a final hit. Chains detected here do not need to be
-accurate.
+colinear and close to each other as a \emph{chain}. We greedily chain the seeds
+while seeding and then filter out short chains that are largely contained in a
+long chain and are much worse than the long chain (by default, both 50\% and
+38bp shorter than the long chain). Chain filtering aims to reduce unsuccessful
+seed extension at a later step. Each chain may not always correspond to a final
+hit. Chains detected here do not need to be accurate.
 
@@
 \caption{Percent mapped reads as a function of the false alignment rate under
-different mapping quality cutoff. $10^6$ pairs of 101bp reads are simulated
-from the human reference genome using wgsim (http://bit.ly/wgsim2) with 1.5\%
-substitution errors and 0.2\% indel variants. The insert size follows a normal
-distribution $N(500,50^2)$. The reads are aligned back to the genome either as
-single end (SE; top panel) or as paired end (PE; bottom panel). GEM is
-configured to allow up to 5 gaps and to output suboptimal alignments (option
-`--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb' for PE). GEM does not compute
-mapping quality. Its mapping quality is estimated with a BWA-like algorithm
-with suboptimal alignments available. Other mappers are run essentially with
-the default setting except for specifying the insert size distribution. The CPU
-time in seconds is shown in the parantheses.}\label{fig:eval}
-\end{figure}.
+different mapping quality cutoff. Alignments with mapping quality 3 or lower
+are excluded. An alignment is \emph{wrong} if after correcting clipping, its
+start position is at least 5bp away from the simulated position. $10^6$ pairs
+of 101bp reads are simulated from the human reference genome using wgsim
+(http://bit.ly/wgsim2) with 1.5\% substitution errors and 0.2\% indel variants.
+The insert size follows a normal distribution $N(500,50^2)$. The reads are
+aligned back to the genome either as single end (SE; top panel) or as paired
+end (PE; bottom panel). GEM is configured to allow up to 5 gaps and to output
+suboptimal alignments (option `--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb'
+for PE). GEM does not compute mapping quality. Its mapping quality is estimated
+with a BWA-like algorithm with suboptimal
+alignments available.  Other mappers are run with the default setting except
+for specifying the insert size distribution. The run time in seconds on a
+single CPU core is shown in the parentheses.}\label{fig:eval}
+\end{figure}
 
@@
 of the insert size distribution from reliable single-end hits. For the
-top 100 hits (by default) of each end, BWA-MEM performs SSE2-based
-Smith-Waterman alignment (SW; \citealt{Farrar:2007hs}) for the mate within a
-window $[\mu-4\sigma,\mu+4\sigma]$ from each hit. The second best SW score is
-recorded to detect potential mismapping in a long tandem repeat. Hits found
-from both the single-sequence alignment and SW rescuing will be used for
-pairing. It is worth noting that BWA and BWA-SW only performs SW for the
-best unique hit, but BWA-MEM also does SW for repetitive hits.
+top 100 hits (by default) of either end, if the mate is unmapped in
+a window $[\mu-4\sigma,\mu+4\sigma]$ from each hit, BWA-MEM performs SSE2-based
+Smith-Waterman alignment (SW; \citealt{Farrar:2007hs}) for the mate within the
+window. The second best SW score is recorded to detect potential mismapping in
+a long tandem repeat. Hits found from both the single-sequence alignment and SW
+rescuing will be used for pairing. %It is worth noting that BWA
+%and BWA-SW only performs SW for the best unique hit, but BWA-MEM also does SW
+%for repetitive hits.
 
@@
 speed, BWA-MEM is similar to GEM and Bowtie2 for this data set, but is about
-6 times as fast as Bowtie2 for a 650bp long-read data set.
+6 times as fast as Bowtie2 and Cushaw2 for a 650bp long-read data set.
 
-To evaluate the performance for even longer query sequences, we aligned the
-{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
-the 536 strain (AC:NC\_008253) with both BWA-MEM and
-nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
-giving 105,505 substitution differences between the strains. BWA-MEM is slower.
-It finished the alignment in 131 seconds, reporting 104,321 substitutions of
-which 102,241 overlapping with the nucmer output. Manual examination of the
-substitutions unique to each aligner reveals that most of them fall in short
-regions with very high divergence. It is unclear which aligner is better. Note
-that although nucmer is faster in the evaluation, only BWA-MEM scales well to
-large genomes.
+We speculate BWA-MEM is more performant for longer reads firstly because of its
+advanced seeding algorithm, which identifies most standing seeds with one pass
+of the read, and secondly because of banded dynamic programming, which
+guarantees linear time complexity in the length of query sequences. BWA-MEM and
+BWA-SW are also able to identify chimeric reads, a crucial feature for contig
+alignment but lacked in most NGS long-read mappers. Our earlier
+paper~\citep{Li:2010fk} discussed this in details.
+
+%To evaluate the performance for even longer query sequences, we aligned the
+%{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
+%the 536 strain (AC:NC\_008253) with both BWA-MEM and
+%nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
+%giving 105,505 substitution differences between the strains. BWA-MEM is slower.
+%It finished the alignment in 131 seconds, reporting 104,321 substitutions of
+%which 102,241 overlapping with the nucmer output. Manual examination of the
+%substitutions unique to each aligner reveals that most of them fall in short
+%regions with very high divergence. It is unclear which aligner is better. Note
+%that although nucmer is faster in the evaluation, only BWA-MEM scales well to
+%large genomes.
 
 BWA-MEM is a fast and accurate aligner for sequence reads and is one of the few
-that work well for both short reads and long sequences up to a few megabases.
+that work well for both 70bp reads and long sequences up to a few megabases.
 Technically, it is possible to make BWA-MEM even faster for long sequences by

### 2013-05-26 bwamem.tex
The version to appear in arXiv as the 2nd version
The supplementary is for a record only. Some of the content has been moved back
to the preprint.

@@
 \centering
-\includegraphics[width=0.4\textwidth]{alnroc-se}\\
-\includegraphics[width=0.4\textwidth]{alnroc-pe}
+\includegraphics[width=0.5\textwidth]{alnroc-color-se}\\
+\includegraphics[width=0.5\textwidth]{alnroc-color-pe}
 \caption{Percent mapped reads as a function of the false alignment rate under
@@
 are excluded. An alignment is \emph{wrong} if after correcting clipping, its
-start position is at least 5bp away from the simulated position. $10^6$ pairs
+start position is within 20bp from the simulated position. $10^6$ pairs
 of 101bp reads are simulated from the human reference genome using wgsim
@@
 for PE). GEM does not compute mapping quality. Its mapping quality is estimated
-with a BWA-like algorithm with suboptimal
-alignments available.  Other mappers are run with the default setting except
-for specifying the insert size distribution. The run time in seconds on a
-single CPU core is shown in the parentheses.}\label{fig:eval}
+with a BWA-like algorithm with suboptimal alignments available. Other mappers
+are run with the default setting except for specifying the insert size
+distribution.
+%GSNAP and STAR essentially do not report mapping quality, so they are not shown in the plot.
+%For SE, GSNAP's accuracy is placed at $(2.8\times10^{-3},96.1\%)$ in the plot and STAR at $(4.6\times10^{-3},95.5\%)$.
+%For PE, GSNAP achieves $(1.1\times10^{-3},97.8\%)$ accuracy and STAR $(2.0\times10^{-3},96.8\%)$.
+The run time in seconds on a single CPU core is shown in the
+parentheses.}\label{fig:eval}
 \end{figure}
@@
 We evaluated the performance of BWA-MEM on simulated data together with
-NovoAlign (http://novocraft.com), GEM, Bowtie2, Cushaw2, BWA-SW and BWA
-(Figure~\ref{fig:eval}). On accuracy, NovoAlign is the best. BWA-MEM is close
-to NovoAlign for PE reads and is comparable to GEM and Cushaw2 for SE.  On
-speed, BWA-MEM is similar to GEM and Bowtie2 for this data set, but is about
+NovoAlign (http://novocraft.com), GEM, Bowtie2, Cushaw2, SeqAlto~\citep{Mu:2012fk}, %GSNAP~\citep{Wu:2010uq}, STAR~\citep{Dobin:2013kx},
+BWA-SW
+and BWA (Figure~\ref{fig:eval}). On accuracy, NovoAlign is the best. BWA-MEM
+is close to NovoAlign for PE reads and is comparable to GEM and Cushaw2 for SE.
+On speed, BWA-MEM is similar to GEM and Bowtie2 for this data set, but is about
 6 times as fast as Bowtie2 and Cushaw2 for a 650bp long-read data set.
@@
 
-%To evaluate the performance for even longer query sequences, we aligned the
-%{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
-%the 536 strain (AC:NC\_008253) with both BWA-MEM and
-%nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
-%giving 105,505 substitution differences between the strains. BWA-MEM is slower.
-%It finished the alignment in 131 seconds, reporting 104,321 substitutions of
-%which 102,241 overlapping with the nucmer output. Manual examination of the
-%substitutions unique to each aligner reveals that most of them fall in short
-%regions with very high divergence. It is unclear which aligner is better. Note
-%that although nucmer is faster in the evaluation, only BWA-MEM scales well to
-%large genomes.
+To evaluate the performance for even longer query sequences, we aligned the
+{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
+the 536 strain (AC:NC\_008253) with both BWA-MEM and
+nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
+giving 105,505 substitution differences between the strains. BWA-MEM is slower.
+It finished the alignment in 131 seconds, reporting 104,321 substitutions of
+which 102,241 overlapping with the nucmer output. Manual examination of the
+substitutions unique to each aligner reveals that most of them fall in short
+regions with very high divergence. It is unclear which aligner is better. Note
+that although nucmer is faster in the evaluation, only BWA-MEM scales well to
+large genomes.
 

### 2013-05-26 mem-supp.tex
The version to appear in arXiv as the 2nd version
The supplementary is for a record only. Some of the content has been moved back
to the preprint.

\documentclass[10pt,pdftex]{article}
\usepackage{graphicx}
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{caption}
\renewcommand{\captionfont}{\fontsize{9pt}{10pt}\selectfont}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}

\addtolength{\textwidth}{3cm}\addtolength{\hoffset}{-1.5cm}\addtolength{\textheight}{4cm}\addtolength{\voffset}{-2cm}

\title{Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM (Supplementary Information)}
\author{Heng Li}
\date{}

\begin{document}

\maketitle

\section{Workflow of BWA-MEM}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{mem-flow}
\caption{BWA-MEM workflow.}
\end{figure}

\section{Evaluation on bacterial genome alignment}
To evaluate the performance for even longer query sequences, we aligned the
{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
the 536 strain (AC:NC\_008253) with both BWA-MEM and
nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
giving 105,505 substitution differences between the strains. BWA-MEM is slower.
It finished the alignment in 131 seconds, reporting 104,321 substitutions of
which 102,241 overlapping with the nucmer output. Manual examination of the
substitutions unique to each aligner reveals that most of them fall in short
regions with very high divergence. It is unclear which aligner is better. Note
that although nucmer is faster in the evaluation, only BWA-MEM scales well to
large genomes.

\section{Evaluation of additional mappers}
In addition to the mappers shown in the main text, we have also evaluated
SeqAlto~\citep{Mu:2012fk}, GSNAP~\citep{Wu:2010uq}, STAR~\citep{Dobin:2013kx},
SMALT (http://bit.ly/smalt-aln) and LAST~\citep{Shrestha:2013aa}. The detailed
command lines are shown in Table S1, S2 and S3.

\begin{table}[!tbhp]
\footnotesize
\centering
\begin{tabular}{lll}
\hline
Program & Version & Indexing command line \\
\hline
Bowtie2 & 2.1.0 & bowtie2-build ref.fa ref.bt2 \\
BWA & 0.7.0 & bwa index ref.fa (version 0.6.2 for BWA and BWA-SW)\\
Cushaw2 & 2.1.8 & cushaw2\_index -a bwtsw ref.fa \\
GEM & 2012-10-28 & gem-indexer -i ref.fa -o ref \\
GSNAP & 2013-03-31 & gmap\_build -d ref.gmap ref.fa \\
LAST & 278 & lastdb -m1111110 ref.last ref.fa \\
Novoalign & 2.08.03 & novoindex -s3 ref.novo ref.fa \\
SeqAlto & 0.5-r123 & seqalto index ref.fa 22 ref.alto \\
SMALT & 0.7.2 & smalt index -k20 -s3 ref.smalt ref.fa \\
STAR & 2.3.0e & STAR --runMode genomeGenerate --genomeDir star --genomeFastaFiles ref.fa \\
\hline
\end{tabular}
\caption{Program versions and indexing command lines}
\end{table}

\begin{table}[!tbhp]
\footnotesize
\centering
\begin{tabular}{lrrp{9cm}}
\hline
Algorithm & Time:s & Mem:GB & Command line \\
\hline
Bowtie2   & 582 & 3.1 & bowtie2 -x ref.bt2 r12.fq \\
BWA       & 1055& 4.5 & bwa aln -f r12.sai ref.fa r12.fq; bwa samse ref.fa r12.sai r12.fq \\
BWA-SW    & 1002& 5.1 & bwa bwasw ref.fa r12.fq \\
BWA-MEM   & 467 & 5.1 & bwa mem ref.fa r12.fq \\
Cushaw2   & 967 & 3.3 & cushaw2 -r ref.fa -q r12.fq \\
GEM       & 426 & 4.0 & gem-mapper -I ref.gem -i r12.fq -e5 -m5 -s1 \\
GSNAP     & 5553& 5.0 & gsnap -Asam -d ref.gmap r12.fq \\
LAST      & 4302& & lastal -Q1 ref.last r12.fq $|$ last-map-probs.py -m 0.5 - $|$ maf-convert.py sam - \\
Novoalign & 3960& 6.0 & novoalign -oSAM -rRandom -d ref.novo -f r12.fq \\
SeqAlto   & 850 & 6.8 & seqalto align ref.alto -1 r12.fq \\
SMALT     & 3859& 3.4 & smalt map -f samsoft ref.smalt r12.fq \\
STAR      & $<$128 & 26.6& STAR --genomeDir star --readFilesIn r12.fq --runThreadN 1 --outFilterMultimapNmax 1 --chimSegmentMin 30 --genomeLoad LoadAndKeep \\
\hline
\end{tabular}
\caption{CPU time in seconds, peak memory in GB and command lines for
single-end mapping.  Programs were run on a single Xeon CPU core clocked at
3.47GHz with 96GB RAM. They were timed for the second run to minimize index
loading time as the system can usually cache the index in memory. However, STAR
still spent significant amount of time for loading the index. The actual CPU
time spent on alignment should be shorter.}
\end{table}

\begin{table}[!tbhp]
\footnotesize
\centering
\begin{tabular}{lrrp{9cm}}
\hline
Algorithm & Time:s & Mem:GB & Command line \\
\hline
Bowtie2   & 545& 3.1& bowtie2 -x ref.bt2 -1 r1.fq -2 r2.fq -X650 \\
BWA       & 1055& 5.1 & bwa aln -f r1.sai ref.fa r1.fq; bwa -f r2.sai ref.fa r2.fq; bwa sampe ref.fa r1.sai r2.sai r1.fq r2.fq \\
BWA-SW    & 1002& 5.1 & bwa bwasw ref.fa r1.fq r2.fq \\
BWA-MEM   & 467 & 5.1 & bwa mem ref.fa r1.fq r2.fq \\
Cushaw2   & 1026& 3.3 & cushaw2 -r ref.fa -q r1.fq r2.fq \\
GEM       & 529 & 4.0 & gem-mapper -I ref.gem -1 r1.fq -2 r2.fq -e5 -m5 -s1 -pb \\
GSNAP     & 3514& 5.0 & gsnap -Asam -d ref.gmap r1.fq r2.fq \\
LAST      & 5386& & lastal -Q1 ref.last r1.fq $>$ r1.maf; lastal -Q1 ref.last r2.fq $>$ r2.maf; last-pair-probs.py -m 0.5 r1.maf r2.maf $|$ maf-convert.py sam - \\
Novoalign & 2585& 6.0 & novoalign -i 500 50 -oSAM -rRandom -d ref.novo -f r1.fq r2.fq \\
SeqAlto   & 879 & 6.8 & seqalto align ref.alto -1 r1.fq -2 r2.fq -i 650 -m 500 \\
SMALT     & 1705& 3.4 & smalt map -f samsoft -i 650 -j 450 ref.smalt r1.fq r2.fq \\
STAR      & $<$135& 26.6& STAR --genomeDir star --readFilesIn r1.fq r2.fq --runThreadN 1 --outFilterMultimapNmax 1 --chimSegmentMin 30 --genomeLoad LoadAndKeep \\
\hline
\end{tabular}
\caption{CPU time in seconds, peak memory in GB and command lines for mapping paired-end reads.}
\end{table}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{alnroc-color-se}\\
\includegraphics[width=\textwidth]{alnroc-color-pe}
\caption{Percent mapped reads as a function of the false alignment rate under
different mapping quality cutoff. STAR does not output mapping quality. Almost
all GSNAP alignments have mapping quality 18. Their accuracy is represented by
a single dot only.}
\end{figure}

\pagebreak
\bibliography{bwamem}

\end{document}

### 2013-05-26 bwamem.tex
removed a vspace{}
better formatting in arXiv; change to abstract

@@
 It automatically chooses between local and end-to-end alignments, supports
-paired-end reads and reports multiple local hits. The algorithm is robust to
+paired-end reads and performs chimeric alignment. The algorithm is robust to
 sequencing errors and applicable to a wide range of sequence lengths from 70bp
@@
 
-In this background, a few long-read alignment algorithms, including
-BWA-SW~\citep{Li:2010fk}, Bowtie2~\citep{Langmead:2012fk},
+In this background, a few long-read alignment algorithms, notably including
+BWA-SW~\citep{Li:2010fk}, Bowtie2 \citep{Langmead:2012fk},
 Cushaw2~\citep{Liu:2012uq} and GEM~\citep{Marco-Sola:2012kx}, have been
@@
 concerns motivated us to explore a new alignment algorithm.
-
-\vspace{-1em}
 
